---
title: "intro-to-computation"
author: "Daniele Melotti"
date: "2023-11-02"
output: pdf_document
---

# 1. Retrieve Data Points

We import a dataset containing 1 variable, which is the age of a set of hypothetical customers, and move the ages into a numeric vector called _ages_.

```{r setup, include=FALSE}
# Loading data
dataset <- read.table("data/customers.txt", header = TRUE, colClasses = "numeric")
ages <- dataset$age
```

Then, we want to determine how many customers (or ages of customers) are present within our data.

```{r}
length(ages)
```

Now we know that there are 399 customers in total. Next, we check what is the age of certain elements of the vector; for instance, the 1st, 5th and 100th element of _ages_.

```{r}
ages[c(1, 5, 100)]
```

We see that the the 1st, 5th and 100th are _49_, _45_, and _34_ years old respectively. Now, we are interested in finding the age of the youngest and oldest customer.

```{r}
min(ages)
max(ages)
```

We can see that the youngest and oldest customers present in the dataset are aged _18_ and _85_ respectively.

# 2. Sorting and High-Level Analysis

We sort the ages into ascending order. We can take a peek into the five lowest ages:

```{r}
sorted_ages <- sort(ages)
sorted_ages[1:5]
```

As we can see, there is only one customer aged 18, while there are at least four customers aged 19. If we were interested in seeing uniquely the 5 smallest ages without caring about how many customers have said ages, we could do it in the following way:

```{r}
head(unique(sorted_ages), n = 5)
```

Hence, we know that the 5 smallest ages are _18_, _19_, _20_, _21_ and _22_. We repeat the process for the 5 highest ages:

```{r}
tail(unique(sorted_ages), n = 5)
```

Next, we compute a statistic of central tendency, the mean:

```{r}
mean(ages)
```

The average age of the customers is 46.807, close to 47. As for the standard deviation:

```{r}
sd(ages)
```

As we can see, the standard deviation is 16.3698, which is roughly a third of the mean.

# 3. Analysis of Differences

# 4. Visualization

We see that the 5 highest ages are _80_, _81_, _82_, _83_ and _85_. In order to get a more holistic view on the distribution of _ages_, we can plot a histogram.

```{r}
hist(ages, main = "Customer Ages Histogram", xlab = "Age", ylab = "Frequency", breaks = 35)
```

